---
title: "Exploring Relationships of Physical and Socioeconomic Factors Associated with Water Temperature."
author: "Cammie Moore, Jack Killeen, Shaochong Xue"
date: "2025-04-04"
output: html_document
editor_options: 
  chunk_output_type: console
---

*Link to GitHub repository:https://github.com/AllenX0627/MooreXueKilleen.git *


```{r # 0. Set up R environment setup, include=FALSE}
rm(list=ls())

library(tidyverse)
library(readxl)
library(readr)
library(here)
library(dataRetrieval)
library(sf)
library(mapview)
library(tidycensus)
library(tigris)
library(prism)
library(raster)
library(GGally)
library(car)
library(ggplot2)
library(kableExtra)


options(scipen = 999)

```



```{r # 1. Get states socioeconomic data census.gov, include=FALSE}

income <- get_acs(
  geography = "county",
  variables = c(total_pop = "B01003_001", med_income = "B19013_001", poverty = "B17001_002", employed_mine = "C24030_003", total_employed = "C24030_001", employed_constr = "C24030_005", employed_manu = "C24030_006"), 
  state = c("NC", "GA", "SC", "TN"),
  year = 2023,
  survey = "acs5", # american community survey 5 year
  geometry = TRUE
)

income <- income|> #pivot wider to get the data out of the variable column
  dplyr::select(!moe)|>
  pivot_wider(names_from = variable, 
              values_from = estimate)
census_tidy <- income|> #finding percentages of values for tidy df
  mutate(
    per_pov = (poverty/total_pop)*100,
    per_industry = ((employed_mine+employed_constr+employed_manu)/total_employed)*100
  )|>
  dplyr::select(GEOID, NAME, total_pop, med_income, per_pov, per_industry)
# above code got percent poverty rates and percent industry rates for each county. "Industry" is defined as someone reporting on the ACS as working in construction, mining, or manufacturing

```



```{r # 2. Get water temperature data for sites sites reading, include=FALSE}
sites.df <- read.delim("Data/site_list.txt", comment.char = "#", stringsAsFactors = FALSE) # the dates of this file is from 2025-04-01 to 2025-04-02

sites.df <- renameNWISColumns(sites.df) # renaming
sites.df$site_no <- as.factor(sites.df$site_no)
sites.df$Temp_C <- as.numeric(sites.df$Temp_C)
sites.df <- drop_na(sites.df)



sites.tidy.df <- sites.df|> #mean water temperature from past seven days for each site
  group_by(site_no)|>
  summarise(mean_temp_C = mean(Temp_C, na.rm=TRUE))|> # water temp in degrees Celsius
  filter(!is.na(mean_temp_C))

list_sites <- as.list(sites.tidy.df$site_no) #list of sites only
site_info <- readNWISsite(unlist(list_sites)) #geo references for list of sites datum NAD83
site_info <- site_info |>
  dplyr::select(site_no, dec_lat_va, dec_long_va)

sites_geo <- left_join(sites.tidy.df, site_info, by = "site_no") #joining sites

sites.shp <- sites_geo |> 
  st_as_sf(
    coords = c('dec_long_va','dec_lat_va'),
    crs=4326
    )

```

```{r writing .shp for gauges, eval=FALSE, include=FALSE} 
st_write( # change eval to TRUE the first time running code to write the file
  sites.shp,
  here("Data/sites.shp"),
  driver='ESRI Shapefile',
  append = FALSE
  )
mapview(sites.shp)
```

### Research Question
How is stream water temperature related to physical and socioeconomic factors such elevation, income, and concentration of heavy industry?


### Exploring Data

```{r setting graph themes, include=FALSE}
mytheme <- theme(
  legend.title = element_text(size = 11),
  legend.position = "bottom"
)
```
Census data was collected from the 2023 ACS 5-year community survey. Socioeconomic factors such as median income, employment, and total population were analyzed to explore their potential relationship with water temperatures recorded at various USGS gauges across North Carolina, Tennessee, South Carolina, and Georgia. Water temperature data were collected from April 1st to April 2nd and averaged to obtain the mean temperature (in Celsius) for that period.

```{r exploring mapping rivers and counties, echo=FALSE, fig.cap= "Figure 1. Exploring River Sites and Percentage of Populations in Industry"}
census_tidy.utm <- st_transform(census_tidy, 32617)
sites.utm <- st_transform(sites.shp, 32617)

rivers_county.plot <- ggplot()+
  geom_sf(data = census_tidy.utm, aes(fill = per_industry))+
  geom_sf(data = sites.utm, aes(color = "River Temperature Sites (C)"))+
  scale_color_manual(name = NULL, values = c("River Temperature Sites (C)" = "blue"))+
  scale_fill_continuous(low = "red", high = "lightyellow", name="Percentage of Employed Persons in Industry", text )+
  mytheme

rivers_county.plot
```
As shown in exploratory Figure 1, there are varying degrees of industry (defined as employment in construction, manufacturing, or mining) across Georgia, North Carolina, South Carolina, and Tennessee. There appears to be a concentration of river gauges in northern Georgia near Atlanta. There may be spatial autocorrelation to take into consideration due to this concentration of data.

```{r # 3. Get other environmental data for sites flow data, include=FALSE}
flow_data <- readNWISdata(
  service = "dv",
  site = sites.tidy.df$site_no,
  parameterCd = "00060",
  startDate = "2025-04-01",
  endDate = "2025-04-02"
)

flow_data <- renameNWISColumns(flow_data)

flow_tidy <- flow_data %>%
  group_by(site_no) %>%
  summarise(mean_flow = mean(Flow, na.rm = TRUE)) %>%
  filter(!is.na(mean_flow))

head(flow_tidy)

```


```{r elevation data, include=FALSE}
# use elevatr package to get elevation data for each sites
library(elevatr)

elev.df <- get_elev_point(sites.utm, src = "aws")
```

```{r use PRISM package to get air temp data, include=FALSE}
options(prism.path = "Data/PRISM")

get_prism_dailys(
  type = "tmean",
  minDate = "2025-04-01",
  maxDate = "2025-04-02",
  keepZip = FALSE
)

```

```{r get tmean for each sites, include=FALSE}
# read PRISM raster file
tmean_0401 <- raster("Data/PRISM/PRISM_tmean_early_4kmD2_20250401_bil/PRISM_tmean_early_4kmD2_20250401_bil.bil")
tmean_0402 <- raster("Data/PRISM/PRISM_tmean_early_4kmD2_20250402_bil/PRISM_tmean_early_4kmD2_20250402_bil.bil")

# get temp for each sites by coordinate 
temp_0401 <- raster::extract(tmean_0401, sites.shp)
temp_0402 <- raster::extract(tmean_0402, sites.shp)

# calculate t mean
temp_mean <- rowMeans(cbind(temp_0401, temp_0402), na.rm = TRUE)
temp_mean

```


```{r # 4. Combine data st join, include=FALSE}
#st join to get pop, income, pov, industry for each sites
sites.county <- st_join(sites.utm, census_tidy.utm)
head(sites.county)
```


```{r combine all data, include=FALSE}
sites.full <- sites.county %>%
  st_drop_geometry() %>%
  dplyr::select(site_no, mean_temp_C, total_pop, med_income, per_pov, per_industry) %>%
  left_join(flow_tidy, by = "site_no") %>%
  mutate(elev = elev.df$elevation,
         mean_t_air = temp_mean
         ) %>%
  drop_na()
  
head(sites.full) # please use this df for analysis!
```

### Data Structure Table

```{r data structure table, fig.cap= "Table 1. Summary of variables used in the analysis", echo=FALSE}
library(knitr)


variables_summary <- data.frame(
  Variable = c(
    "mean_temp_C", "mean_t_air", "mean_flow", "log_flow",
    "elevation", "log_elev", "total_pop", "med_income",
    "per_pov", "per_industry"
  ),
  Unit = c(
    "°C", "°C", "cfs", "log(cfs)", "meters", "log(meters)",
    "people", "USD", "%", "%"
  ),
  Range = c(
    paste0(round(min(sites.full$mean_temp_C),1), "–", round(max(sites.full$mean_temp_C),1)),
    paste0(round(min(sites.full$mean_t_air),1), "–", round(max(sites.full$mean_t_air),1)),
    paste0(round(min(sites.full$mean_flow),1), "–", round(max(sites.full$mean_flow),1)),
    paste0(round(min(log(sites.full$mean_flow)),1), "–", round(max(log(sites.full$mean_flow)),1)),
    paste0(round(min(sites.full$elev),1), "–", round(max(sites.full$elev),1)),
    paste0(round(min(log(sites.full$elev)),1), "–", round(max(log(sites.full$elev)),1)),
    paste0(round(min(sites.full$total_pop),0), "–", round(max(sites.full$total_pop),0)),
    paste0(round(min(sites.full$med_income),0), "–", round(max(sites.full$med_income),0)),
    paste0(round(min(sites.full$per_pov),1), "–", round(max(sites.full$per_pov),1)),
    paste0(round(min(sites.full$per_industry),1), "–", round(max(sites.full$per_industry),1))
  ),
  Mean = c(
    round(mean(sites.full$mean_temp_C),1),
    round(mean(sites.full$mean_t_air),1),
    round(mean(sites.full$mean_flow),1),
    round(mean(log(sites.full$mean_flow)),1),
    round(mean(sites.full$elev),1),
    round(mean(log(sites.full$elev)),1),
    round(mean(sites.full$total_pop),0),
    round(mean(sites.full$med_income),0),
    round(mean(sites.full$per_pov),1),
    round(mean(sites.full$per_industry),1)
  ),
  Source = c(
    "USGS stream data", "PRISM climate data", "USGS stream data",
    "Calculated", "AWS Elevation API", "Calculated",
    "US Census ACS", "US Census ACS", "US Census ACS", "US Census ACS"
  )
)


kable(variables_summary, caption = "Table 1. Summary of Variables Used in the Analysis") 

```


### Linear Regressions

```{r scatter_matrix, fig.cap="Figure 2. Scatterplot matrix of water temperature and predictors", echo=FALSE}
# 5. Linear Regressios
sites_scatterplot <- sites.full %>%
  dplyr::select(-site_no)

ggpairs(sites_scatterplot, axisLabels = "show")
```

As shown in Figure.2:

Air temperature is positively correlated with water temperature (r = 0.635).

Elevation is negatively correlated with water temperature (r = -0.646).

Poverty rate shows a weak positive correlation with water temperature (r = 0.247).

Other variables, such as flow, industrial employment percentage, and median income, do not show significant correlations with water temperature. 

Elevation and flow are highly skewed, we applied log transformations to improve linearity for regression.

```{r log transformation, include=FALSE}
sites.full.log <- sites.full %>%
  mutate(
    log_flow = log(mean_flow),
    log_elev = log(elev)
  )
head(sites.full.log)
```

```{r scatter plot matrix 2, fig.cap="Figure 3. Scatterplot matrix of water temperature and predictors (log transformed)", echo=FALSE}
sites_scatterplot_2 <- sites.full.log %>%
  dplyr::select(-site_no, -mean_flow, -elev)

ggpairs(sites_scatterplot_2, axisLabels = "show")
```

AS shown in Figure 3, after applying log transformations to elevation and flow, the overall patterns in the scatterplot matrix did not change much.

```{r linear model, include=FALSE}
model_1 <- lm(mean_temp_C ~ mean_t_air + log_flow + log_elev + per_pov + per_industry + med_income, 
                 data = sites.full.log)
summary(model_1)
```

According to the linear model, air temperature, stream flow, elevation, and poverty rate were significant predictors of water temperature.

Every 1°C increase in air temperature is associated with 0.35°C increased in water temperature (p = 0.0124).

Every 1% increase in the flow is associated with 0.0034°C decrease in water temperature (p = 0.00019).

Every 1% increase in the flow is associated with 0.0092°C decrease in water temperature (p = 0.0055).

Every 1% increase in the poverty rate is associated with 0.18°C increase in water temperature (p = 0.0213).

Industrial employment and median income were not significantly related to water temperature.

The model explained about 48% of the variation in water temperature across our study sites (Adjusted R² = 0.478), suggesting a moderately strong fit.

```{r VIF, include=FALSE}
vif(model_1)
```

VIF values for all predictors were below 5, indicating no serious multicollinearity concerns in the model.

```{r AIC to select variables, include=FALSE}
model_2 <- step(model_1, direction = "both")

summary(model_2)

```

```{r AIC comparison, include=FALSE}
AIC(model_1, model_2)
```

We used AIC to select variables, reducing the linear regression model.The final model included air temperature, flow, elevation, and poverty rate as predictors.

The reduced model has a slightly higher Adjusted R² (0.480 vs 0.478) and lower AIC (514 vs 517), indicating a better fit with fewer predictors.

### Difference Test

Based on the U.S. Census Bureau (2023) report, the median household income in the United States was $74.580.

Therefore, to further explore the relationship between income levels and water temperature, we classified our data into two groups: "High Income" (median income ≥ 74.580) and "Low Income" (median income < 74.580).


```{r create income groups, include=FALSE}
sites.full.t <- sites.full.log %>%
  mutate(income_group = ifelse(med_income < 74580, "Low", "High")) %>%
  mutate(income_group = factor(income_group))

head(sites.full.t)
```

```{r boxplot, fig.cap="Figure 4: Stream Water Temperature by Income Group",  echo=FALSE}
boxplot(mean_temp_C ~ income_group,
        data = sites.full.t,
        xlab = "Income Group",
        ylab = "Mean Water Temperature (°C)",
        col = c("navy", "orange"))
```

```{r normality test, include=FALSE}
shapiro.test(sites.full.t$mean_temp_C[sites.full.t$income_group == "Low"])
shapiro.test(sites.full.t$mean_temp_C[sites.full.t$income_group == "High"])

```

Based on the Shapiro-Wilk test results, water temperature data for the high income group significantly deviates from a normal distribution. Therefore, we used the Wilcoxon test to compare water temperatures between the two income groups.

```{r wilcoxon difference test, include=FALSE}
wilcox.test(mean_temp_C ~ income_group, data = sites.full.t)
```

The result of Wilcoxon test was not statistically significant (p = 0.903), indicating that water temperature did not differ significantly between high-income and low-income sites.

### Summary and Conclusions

Our analysis shows that air temperature, elevation, stream flow, and poverty rate are significant predictors of river water temperature in the southeastern United States. Specifically, water temperature increases with higher air temperature and poverty rate, and decreases with higher elevation and stream flow. However, when comparing sites grouped by income level, no statistically significant difference in water temperature was found.

While industrial activity is commonly associated with thermal pollution and elevated air temperatures, our study did not find its significant relationship with water temperature. This may be due to several reasons:

1. The industrial employment variable used was broad and may not reflect actual thermal discharges.

2. Most sites likely comply with environmental regulations that limit heated water discharge.

3. Stream temperature is strongly influenced by physical factors such as air temperature and elevation, which may overshadow weaker socioeconomic effects.

This study has several limitations. In the linear regression models, dominant environmental variables such as air temperature and elevation may obscure the effects of socioeconomic factors. Additionally, high-elevation or mountainous regions often tend to be lower-income, introducing potential multicollinearity among predictors. These relationships should be examined more carefully in future work, ideally with larger datasets and methods accounting for spatial dependence.


#### Data Citation
Oliver, S.K., Appling, A., Watkins, D., Atshan, R., and Read, J., 2024, Compilation of multi-agency water temperature observations for U.S. streams, 1894-2022: U.S. Geological Survey data release, https://doi.org/10.5066/P9EMWZ35.



